Manual classification of tumor cells

### 2024-05-14 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
add template notebooks

---
title: "Manual classification of tumor cells"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

@@

# output tumor/normal classifications
results_dir <- file.path(module_base, "results", "marker_gene_analysis")
fs::dir_create(results_dir)

classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
output_classifications_file <- file.path(results_dir, classifications_filename)
```

@@

@@

Looking at this plot, there is definitely some variation in the expression of marker genes. 
It looks like the highest gene expression is seen in the larger group of cells on the bottom right of the UMAP.

We can also look at the distributions for each marker gene. 
I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.

@@

Although we see some slight semblance of a bimodal distribution for most marker genes, it is hard to see and we cannot directly compare gene expression values across genes. 
This would make it hard to identify a cut off to categorize cells as expression or not expressing the marker gene.

Now we will transform each of the gene expression vectors by generating z-scores. 
Then we might be able to find a cut off we can use across samples for if marker genes are present in a cell or not.

@@

Now we can create the same density plot but looking at z-scores.

@@

It looks like some marker genes have distinct groups of cells with z-score > 0, while other marker genes may not be as informative. 
Keep in mind this could vary somewhat across patients, so I'm wary to remove marker genes without looking across multiple references.

@@

Let's try and use the marker gene expression to classify tumor cells. 
It looks like we could use a cutoff of z-score > 0 to count that cell as a tumor cell.

We could either count any cell that expresses at least one marker gene > 0 as a tumor cell, or look at the combined expression. 
Let's start with classifying tumor cells as tumor if any marker gene is present (z-score > 0). 
This also means the sum of all z-scores > 0.

@@

# plot mean gene expression 
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c()
```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

@@

It looks like ADGRG2, HES1, NROB1, and STEAP1 don't show a lot of difference between cells that might be tumor vs. normal cells and are generally lowly expressed.
CD99 appears to be expressed in both normal and tumor at similar levels. 
All other genes show differences between the two, although at varying degrees. 
Again, we should make sure this is consistent across multiple patients before removing these genes as marker genes.

@@

Below, we will use `scran::findMarkers()` to find any markers between tumor and normal cells. 
Since we used marker genes to classify tumor cells, I would expect that at least some of the top markers would be from our list of tumor markers.

@@

We would expect that cells we are labeling as tumor cells have expression of the tumor marker genes. 
Cells labeled as normal should not have expression of the markers.

@@

Looking at this might help give us an idea of genes that are more specifically expressed in tumor cells. 
Genes like ADGRG2 and HES1 do not seem specific to tumor cells. 
There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.

CD99, CAV1, and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few normal cells. 
This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.

KDSR, MAPT, NKX2-2, PAPPA, and PRKCB all seem to have a high proportion of tumor cells expressing the gene and a low proportion of normal cells that express the gene. 
PRKCB and MAPT are also 2 of the top genes identified using the marker analysis.

Almost all genes have at least some expression in normal cells. 
Perhaps those cells should also be categorized as tumor cells, or they just have some baseline expression in normal cells which is possible.

@@

The last thing we will do here is compare our manual annotations with the annotations using `SingleR` and `CellAssign`. 
Looking at the annotations and UMAPs in the supplemental report, it looks like our group of "Tumor" cells are classified mostly as "Muscle cells" by both SingleR and CellAssign. 
This makes sense since Ewing's sarcoma cells tend to resemble muscle cells.

I would expect the Endothelial cells, Fibroblasts, and Macrophages assigned by `CellAssign` and `SingleR` to line up with the normal cells in this sample.

@@

When looking at `SingleR`, we see that the tumor cells line up with smooth muscle cells, chondrocytes, neurons, and fibroblasts. 
Ewing's cells have a neuronal like phenotype and express high levels of ECM genes, which may explain the overlap with chondrocytes and fibroblasts. 
We do see that macrophages and endothelial cells tend to line up with the normal cells, which is expected.

For CellAssign, all tumor cells are pulmonary vascular smooth muscle cells. 
We also see tha normal cells contain some smooth muscle cells, but mostly are fibroblasts and endothelial cells.

Looking at the heatmap in the supplemental report, the endothelial cells called by SingleR and CellAssign have a lot of overlap. 
Let's look at expression of marker genes in these endothelial cells compared to cells that are categorized as smooth muscle cells.

```{r}
select_celltype_df <- celltypes_df |> 
  dplyr::mutate(
    automated_annotation = dplyr::case_when(
      singler_celltype_annotation == "smooth muscle cell" & cellassign_celltype_annotation == "Pulmonary vascular smooth muscle cells" ~ "Muscle cell",
      singler_celltype_annotation == "endothelial cell" & cellassign_celltype_annotation == "Endothelial cells" ~ "Endothelial cell"
    )
  ) |> 
  dplyr::select(barcodes, automated_annotation)

annotation_plot_df <- umap_df |> 
  dplyr::left_join(select_celltype_df) |>
  dplyr::filter(!is.na(automated_annotation))

ggplot(annotation_plot_df, aes(x = transformed_gene_expression, color = automated_annotation)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) 
```

Consistent with previous findings, ADGRG2, HES1, NROB1, and STEAP1 don't seem to show any expression in either groups of cells. 
CCND1 and CD99 appear to have similar expression across both groups of cells, while MAPT and PRKCB seem to have the most distinct expression profiles between muscle and endothelial cells.

## Conclusions

- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here. - Using a z-score cut-off identifies a group of cells that appear to be tumor. .
- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
- It's likely that ADGRG2, HES1, NROB1, and STEAP1 will not be informative in assigning tumor cells. 
- We may be weary of genes like CD99 too as expression is more consistent across cells rather than specific to a group of cells. 
- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

>>>>

---
title: "Manual classification of tumor cells"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
  results_dir: "../results/annotate_tumor_cells_output"
  reference_cell_file: "../references/cell_lists/SCPCS000490/SCPCL000822_reference-cells.tsv"
---

@@

```{r}
project_root <- here::here()
renv::load(project_root)
```

@@

# output tumor/normal classifications
fs::dir_create(params$results_dir)

classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
output_classifications_file <- file.path(params$results_dir, classifications_filename)
```

@@

```{r}
# read in normal and tumor cells for later comparisons 
if(file.exists(params$reference_cell_file)){
  ref_cells_df <- readr::read_tsv(params$reference_cell_file)
} else {
  ref_cells_df <- NULL 
}
```

@@

We can also look at the distributions for each marker gene. 
We would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.

@@

Now we will transform each of the gene expression vectors by generating z-scores and use a cutoff of z-score > 0 to call tumor cells.

@@

@@

@@

To classify tumor cells, we use a cutoff of z-score > 0 to count that cell as a tumor cell. 
We will call cells as tumor if any marker gene is present (z-score > 0). 
This also means the sum of all z-scores > 0.

@@

# plot mean gene expression 
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c()
```

@@

@@

Below, we will use `scran::findMarkers()` to find any markers between tumor and normal cells. 
Since we used marker genes to classify tumor cells, we expect that at least some of the top markers would be from our list of tumor markers.

@@

We expect that cells we are labeling as tumor cells have expression of the tumor marker genes. 
Cells labeled as normal should not have expression of the markers.

@@

Looking at this might help give us an idea of genes that are specifically expressed in only tumor cells.

@@

The last thing we will do here is compare the manual annotations with the annotations from `SingleR` and `CellAssign`.

@@

```{r}
# plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign 
# use already identified normal/tumor cell types 
if(!is.null(ref_cells_df)){

  # get list of normal cells 
  normal_cells <- ref_cells_df |> 
    dplyr::filter(reference_cell_class == "Normal") |> 
    dplyr::pull(barcodes) |> 
    unique()

  # get list of tumor cells 
  tumor_cells <- ref_cells_df |> 
    dplyr::filter(reference_cell_class == "Tumor") |> 
    dplyr::pull(barcodes) |> 
    unique()

  # only create this plot if normal and tumor cells are provided 
  if(length(normal_cells) > 0 && length(tumor_cells) > 0){

    # indicate cells that are part of normal and tumor lists
    select_celltype_df <- celltypes_df |> 
      dplyr::mutate(
        automated_annotation = dplyr::case_when(
          barcodes %in% normal_cells ~ "Normal cell",
          barcodes %in% tumor_cells ~ "Tumor cell"
        )
      ) |> 
      dplyr::select(barcodes, automated_annotation)

    # join with umap df
    annotation_plot_df <- umap_df |> 
      dplyr::left_join(select_celltype_df) |>
      dplyr::filter(!is.na(automated_annotation))

   ggplot(annotation_plot_df, aes(x = transformed_gene_expression, color = automated_annotation)) +
      geom_density() +
      facet_wrap(vars(gene_symbol)) 
  } 
} 
```
